---
title: "Scientific Notation"
format: 
  html:
    code-fold: false
editor: visual
---

Scientific notation is used to express very large or very small numbers in a concise format using exponents of 10 ($10^n$). So, rather than having to show all of the digits for $4,600,000,000$ (4.6 Trillion) it can be expressed in scientific notation: $4.6 \times 10^9$. Likewise, very small numbers can be expressed using the same format, e.g. $0.0000005$ = $5.0 \times 10^{-7}$.

**Note**: for numbers larger than one the exponent is positive ($10^9$) and for numbers less than one the exponent is negative ($10^{-7}$)

**e values** are used to express scientific notation within R (and other programming languages) and essentially the $\text{e}$ replaces the $\times 10$ part of the notation.\
For example, $3.1\text{e}3$ is the same as $3.1 \times 10^3$ (which is the same as 3100):

```{r}
3.1e3 == 3.1 * 10^3
```

Likewise, $2.5\text{e-}3$ is the same as $2.5 \times 10^{-3}$ (which is the same as .0025):

```{r}
2.5e-3 == 2.5 * 10^(-3)
```

However, if you would like to turn off scientific notation in R you can type:

```{r}
options(scipen=999)
```

{{< include eNumbersQuestions.qmd >}}

{{< include ../feedback.qmd >}}
