---
title: "Normal Distribution (incomplete)"
format: html
editor: visual
---

**Parametric statistics** often compare values to a **normal** distribution of **expected results**, based on the **estimated mean and SD**. Lets start by showing a (made up) **normal** distribution of heights in centimeters:

So lets say the average person's height is 150cm, and the standard deviation of height across the population is 10cm. The data would look something like:

```{r}
# Plot a normal distribution of heights
population_heights_x <- seq(
  120,    # min
  180,    # max
  by = 1  
)
population_heights_y <- dnorm(
  population_heights_x,
  mean = 150,
  sd   = 10
)
plot(
  population_heights_x,
  population_heights_y
)
# Add line to show mean
abline(
  v=150,  # where the line for the mean will be 
  lwd=5
)
```



// explain that a z-score distribution is just a standardised version of the above distribution

## z-distribution

```{r}
# Plot a normal distribution of heights
z_score_x <- seq(
  -3,    # min
  3,    # max
  by = .1  
)
z_score_y <- dnorm(
  z_score_x,
  mean = 0,
  sd   = 1
)
plot(
  z_score_x,
  z_score_y
)
```

For example, lets imagine that a man called John thinks he's either unusually tall, or unusally short, but he isn't sure which. He could compare his height to a normal distribution of heights, and see whether he's in the top or bottom 2.5%.

```{r}
# Plot a normal distribution of heights
z_score_x <- seq(
  -3,    # min
  3,    # max
  by = .1  
)
z_score_y <- dnorm(
  z_score_x,
  mean = 0,
  sd   = 1
)
plot(
  z_score_x,
  z_score_y
)
# Add line to show mean
abline(
  v=0,  # where the line for the mean will be 
  lwd=3
)
# Add line to show lower threshold
abline(
  v= -1.96,
  lwd=3,
  lty=3
)
# Add line to show higher threshold
abline(
  v= 1.96,
  lwd=3,
  lty=3
)

```

# draw red blocks to highlight the values less than -1.96 and greater than 1.96

So lets say he's 170cm tall, the average person is 150cm, and the standard deviation of height across the population is 10cm. The data would look something like:

```{r}
# Plot a normal distribution of heights
population_heights_x <- seq(
  140,    # min
  200,    # max
  by = 1  
)
population_heights_y <- dnorm(
  population_heights_x,
  mean = 150,
  sd   = 10
)
plot(
  population_heights_x,
  population_heights_y
)
# Add line to show mean
abline(
  v=150,  # where the line for the mean will be 
  lwd=5
)
# Add line to show John's height
abline(
  v=170, 
  lwd=3,
  lty=3
)
```

*The thick line refers to the mean height, the dashed line refers to John's height*

You can see above that John is pretty unusually tall. To work out whether he's in the top 5%, lets work out his z-score:

```{r}
john_height <- 170
mean_height <- 150
height_sd   <- 10
john_z_score <- (john_height - mean_height)/height_sd
john_z_score
```

So based on the above, John's Z-score is greater than 1.96, which I claimed above was meaningful. To understand why this is meaningful, lets start by plotting John's score as a z-score:

```{r}
# Plot a normal distribution of heights
z_score_x <- seq(
  -1,    # min
  5,    # max
  by = .1  
)
z_score_y <- dnorm(
  z_score_x,
  mean = 0,
  sd   = 1
)
plot(
  z_score_x,
  z_score_y
)
# Add line to show mean
abline(
  v=0,  # where the line for the mean will be 
  lwd=3
)
# Add line to show John's height
abline(
  v=john_z_score,
  lwd=3,
  lty=3
)

```

As you may have noticed, these figures are identical in their distributions and the relative locations of the vertical lines. So when you have calculated a **z-score** you have calculated a standardised value of how many standard deviations you are above (or below) the mean. People have previously calculated that if you get a z-value of 1.96 and above then you are in the top 5 percentile. However, lets double check that using the pnorm function:

```{r}
# compare the values below with the plot of the z_score plot
pnorm(
  q = 1.96, # put in the z-value here
  mean = 0,
  sd = 1,
  lower.tail = F
)

```

The above p-value of .024998 shows that any Z-score above

Note that how you ask a question **greatly** impacts your approach, and so my claim that I'm in the top 5% of the population is very different to a more general claim I'm taller than the average person. So whether I'm in the top 5% of the population is a very
