---
title: "Skewness"
format: 
  html:  
    code-fold: true
editor: visual
---

**Parametric** analyses are based on the assumption that the data you are analysing is normally distributed (see below):

```{r warning=FALSE}
library(ggplot2)
# https://stackoverflow.com/a/12429538
norm_x<-seq(-4,4,0.001)

norm_data_frame<-data.frame(x=norm_x,y=dnorm(norm_x,0,1))

p<- ggplot(
  data = norm_data_frame, 
  aes(
    x = x,
    y = y
  )
) + geom_line()

 p +
   xlim(c(-3,3)) +
   geom_col(aes(fill=x)) +
   scale_fill_gradient2(low = "red", 
                       high = "blue",
                       mid =  "white",
                       midpoint = median(norm_data_frame$x))+
   
   annotate("text", x=-2.3, y=0.01, label= "13.6%") + 
   annotate("text", x=-1.5, y=0.01, label= "34.1%") + 
   annotate("text", x=-0.6, y=0.01, label= "50%") + 
   annotate("text", x=0.5, y=0.01, label= "84.1%") + 
   annotate("text", x=1.5, y=0.01, label= "97.7%") + 
   annotate("text", x=2.25, y=0.01, label= "100%") +
   geom_vline(xintercept = -2) +
   geom_vline(xintercept = -1) +
   geom_vline(xintercept = 0) +
   geom_vline(xintercept = 1) +
   geom_vline(xintercept = 2) +
   xlab("Z-score") +
   ylab("Frequency") +
   theme(legend.position="none")
```

*White in the above figure represents the median. Note that the mean and median overlap in a normal distribution.*

If your data fits a normal distribution, then you can draw conclusions based on certain facts about this distribution, e.g. the fact that 97.7% of your population *should* have a score that is more negative than +2 standard deviations above the mean (*because Z-scores represent standard deviations from the mean*). As a result, if your data is skewed:

```{r warning=FALSE, message=FALSE}
library(fGarch)

skew_data_frame<-data.frame(
  x = norm_x,
  y = dsnorm(
    x = norm_x,
    mean = 0,
    sd = 1,
    xi = 2,
  )
)

ggplot(
  data = skew_data_frame, 
  aes(
    x = x,
    y = y
  )
) + 
  geom_line() + 
  geom_col(aes(fill=x)) +
  scale_fill_gradient2(low = "red", 
                       high = "blue",
                       mid =  "white",
                       midpoint = -.15)+
   theme(legend.position="none") +
  xlim(-3,3) +
  ylab("Frequency") +
  xlab("Z-score")


```

*White represents the median in the figure above.*

As you can see from the above skewed distribution, the median is below the mean, consistent with the data being skewed. Importantly, the assumptions that we can make about what proportion of the population are 1 standard deviation above and below the mean are no longer valid, as more than half the population are below the mean in this case. This would suggest that non-parametric analyses could be more appropriate if your data is skewed.

So now that we know what skewed distributions look like, we now need to **quantify** how much of a problem with skewness there is.

The next section is a breakdown of the formula for those interested in it (but this is *not crucial*).

#### Optional

If we want to manually calculate skewness:

$$
\tilde{\mu_{3}} = \sum((\frac{x_i- \bar{x}{}} {\sigma} )^3) * \frac{N}{(N-1) * (N-2)}
$$ To do this in R you *could* calculate it manually. We'll use Gapminder's data from 2007:

```{r}
library(gapminder)

# create a new data frame that only focuses on data from 2007
gapminder_2007 <- subset(
  gapminder,   # the data set
  year == 2007     
)

hist(gapminder_2007$lifeExp)

positive_skew_n = length(gapminder_2007$lifeExp)
positive_skewness = sum(((gapminder_2007$lifeExp - mean(gapminder_2007$lifeExp))/sd(gapminder_2007$lifeExp))^3) * (
  positive_skew_n / ((positive_skew_n-1) * (positive_skew_n - 2))
)
# to show the output:
positive_skewness
```

... or just


Use code from <https://stackoverflow.com/a/54369572> to give you values for skewness (this has been chosen as this gives skewness and its standard error as calculated by major software like SPSS and JASP):

```{r}

# Skewness and kurtosis and their standard errors as implement by SPSS
#
# Reference: pp 451-452 of
# http://support.spss.com/ProductsExt/SPSS/Documentation/Manuals/16.0/SPSS 16.0 Algorithms.pdf
# 
# See also: Suggestion for Using Powerful and Informative Tests of Normality,
# Ralph B. D'Agostino, Albert Belanger, Ralph B. D'Agostino, Jr.,
# The American Statistician, Vol. 44, No. 4 (Nov., 1990), pp. 316-321

spssSkewKurtosis=function(x) {
  w=length(x)
  m1=mean(x)
  m2=sum((x-m1)^2)
  m3=sum((x-m1)^3)
  m4=sum((x-m1)^4)
  s1=sd(x)
  skew=w*m3/(w-1)/(w-2)/s1^3
  sdskew=sqrt( 6*w*(w-1) / ((w-2)*(w+1)*(w+3)) )
  kurtosis=(w*(w+1)*m4 - 3*m2^2*(w-1)) / ((w-1)*(w-2)*(w-3)*s1^4)
  sdkurtosis=sqrt( 4*(w^2-1) * sdskew^2 / ((w-3)*(w+5)) )
  
  ## z-scores added by reading-psych
  zskew = skew/sdskew
  zkurtosis = kurtosis/sdkurtosis

  mat=matrix(c(skew,kurtosis, sdskew,sdkurtosis, zskew, zkurtosis), 2,
        dimnames=list(c("skew","kurtosis"), c("estimate","se","zScore")))
  return(mat)
}
spssSkewKurtosis(gapminder_2007$lifeExp)

```

The above function generated three columns: $estimate$ of skewness (and kurtosis, but we'll deal with <a href="kurtosis.html" target="_blank">kurtosis</a> separately, $standard$ $error$ (SE) and the $skewness_z$ score. **Z-scores** can capture how unlikely the $skewness$ estimate is considering what you would normally expect. In this case, when you divide skewness by standard error, you get a z-score, and if the **absolute value** (i.e. ignoring whether it is positive or negative) of the z-score is greater than 1.96 then it is significantly skewed. If you want to understand why 1.96 is the main number, check out the subsection on <a href="normal.html" target="_blank">normal distributions</a>.

One way to deal with skewed data is to <a href="transform.html" target="_blank">transform</a> the data.
