---
title: "Central Tendency"
format: html
editor: visual
---

Central tendancy describes typical values of a variable, such as it's mean and median.

### Mean vs. Median vs. Mode

The **mean** is often called the "average" informally, but is actually a specific type of "average". The mean is the average you get when add together a group of numbers, and then divide by the number of items you combined. For example, to calculate the mean life expectancy of countries in 2007, we'll use <a href="https://www.gapminder.org/">gapminder data</a>:
::: panel-tabset
## R
```{r}
# install (if required) and load the gapminder data
if(!require(gapminder)){install.packages("gapminder")}
library(gapminder)

# create a new data frame that only focuses on data from 2007
gapminder_2007 <- subset(
  gapminder,   # the data set
  year == 2007     
)

# a reminder of the data frame
rmarkdown::paged_table(head(gapminder_2007))                 

# total of all years
sum_life_expectancy  = sum(gapminder_2007$lifeExp)

# count the people
n_life_expectancy    = length(gapminder_2007$lifeExp)        
mean_life_expectancy = sum_life_expectancy/n_life_expectancy 
mean_life_expectancy
```

## Python
```{python}
# load the gapminder module and import the gapminder dataset
from gapminder import gapminder

# create a new data frame that only focuses on data from 2007
gapminder_2007 = gapminder.loc[gapminder['year'] == 2007]

# a reminder of the data frame
gapminder_2007

# total of all years
sum_life_expectancy  = gapminder_2007['lifeExp'].sum()

# count the people
n_life_expectancy    = gapminder_2007['lifeExp'].count() 

# calculate mean life expectancy
mean_life_expectancy = sum_life_expectancy/n_life_expectancy 
mean_life_expectancy
```
:::
For those of you who like to double check these things (which is a good instinct), lets see what number you get if you use the r function for mean:

```{r}
mean(gapminder_2007$lifeExp)
```

#### Python

```{python}
gapminder_2007['lifeExp'].mean()
```

*Whew* - it's the same as the manual calculation above.

Now **median** is less known than **mean**, but median is the value in the middle once you order all your data. It's well explained in the first paragraph on wikipedia: <a href="https://en.wikipedia.org/wiki/Median" target="_blank">https://en.wikipedia.org/wiki/Median</a>, so I would suggest looking there. As you can see below, the mean and median are not always the same (in fact, they are usually at least slightly different):

```{r}
mean(gapminder_2007$lifeExp)
median(gapminder_2007$lifeExp)
```

#### Python

```{python}
gapminder_2007['lifeExp'].mean()
gapminder_2007['lifeExp'].median()
```

Finally, the **mode** is simply the most frequent number in your data. So lets now see if the mode is closer to the mean or median:

```{r}
mean(gapminder_2007$lifeExp)
median(gapminder_2007$lifeExp)

# Solution for calculating mode found at https://stackoverflow.com/a/2547918 as there doesn't seem to be a native function:
library(modeest)
mlv(gapminder_2007$lifeExp, method = "mfv")

```

#### Python

```{python}
gapminder_2007['lifeExp'].mode()
```

The mode for this data was actually **every** value **perhaps because** each value was unique! Lets double check that:

```{r}
length(gapminder_2007$lifeExp)
length(unique(gapminder_2007$lifeExp))
```

#### Python

```{python}
# count number of elements in the dataset
gapminder_2007['lifeExp'].count()

# create a vector with the unique values presented in the dataset
num_values = gapminder_2007['lifeExp'].unique()

# get the lenght of the vector
len(num_values)
```

The length of the whole **vector** and the unique values of the vector is the same, confirming that there's no repetition in this data (and so **no** number is the mode). Lets make up some data so that we can look at what the mode is:

```{r}
mode_example_vector <- c(1,2,2,3,4,4)
mlv(mode_example_vector, method = "mfv")
```

#### Python

```{python}
import pandas as pd

# create a vector
mode_example_vector = [1,2,2,3,4,4]

# convert the vector to a pandas dataframe
mode_example_vector = pd.DataFrame(mode_example_vector)

# get the mode
mode_example_vector.mode()
```

In the above data, there are 2 modes, as the numbers 2 and 4 occur the most.
