---
title: "Dispersion (incomplete)"
format: html
editor: visual
---
To understand distribution (later), it's helpful to clarify some more basic concepts around how data is **dispersed** or spread. 

## Range
**Range** simply captures the min and the maximum values. Lets look at the min and max for the life expectancy data from 2007:

```{r}
# load the gapminder data
library(gapminder)

# create a new data frame that only focuses on data from 2007
gapminder_2007 <- subset(
  gapminder,   # the data set
  year == 2007     
)

min(gapminder_2007$lifeExp)
max(gapminder_2007$lifeExp)
```
So the **range** for life expectancy in 2007 was between 39.613 and 82.603.

## Variance
### Population Variance
**Variance** is how much the data varies around a mean. To capture this, we compare each individual's score with the mean, so lets do this with our gapminder data's life expectancy:

```{r}
life_expectancy_variance_table <- data.frame(
  life_expectancy = gapminder_2007$lifeExp,
  diff_from_mean  = gapminder_2007$lifeExp - mean(gapminder_2007$lifeExp)
)

rmarkdown::paged_table(life_expectancy_variance_table)
```
So we know for **each** country how different their life expectacy is to the mean life expectancy. But ideally we would like a single value to summarise variance. Lets see what would happen if we tried to summarise these differences from the mean by calculating the mean difference from the mean:

```{r}
mean(life_expectancy_variance_table$diff_from_mean) 
```

We get a number that is pretty close to zero. 


**e values** can be used to indicate whether the number is multiplied by



there are *before* or *after* a decimal. If there is a "-" symbol, then that states that the 0s are *after* a decimal, like the number above. To illustrate this, lets compare 5.153937e-15 to .000000000000005153937:

```{r}
5.153937e-15 == .000000000000005153937
```

e-15 indicates that there are **15** 




)
e-15 represents that this value if 15 positions after the decimal point. 

The formula for the population's variance (Ïƒ^2) (i.e. when you have everyone from the group you are measuring):

$$
\sigma^2 = \frac{\sum((x_i- \bar{x}{} )^2)} {N}
$$




### Sample variance

## Standard deviation
### Population Standard deviation


```{r}
mean(abs(life_expectancy_variance_table$diff_from_mean))

sqrt(sum(life_expectancy_variance_table$diff_from_mean^2)/(length(life_expectancy_variance_table$diff_from_mean))-1)

sd(life_expectancy_variance_table$life_expectancy)

```

### Sample Standard deviation


## degrees of freedom

The formula for the population's variance (Ïƒ^2) (i.e. when you have everyone from the group you are measuring):

$$
\sigma^2 = \frac{\sum((x_i- \bar{x}{} )^2)} {N}
$$

Ïƒ2 =	Î£(ð‘¥i - xÌ„)2
n
Population standard deviation formula

Ïƒ = âˆš(population variance)
